---
title: "RLB database loader"
output: html_notebook
---
Load the libraries and functions.
```{r}
source("functions.R")
```
Connect to the database.
```{r}
PostgreSQL(max.con = 16,fetch.default.rec = 500,force.reload = FALSE)
driver<-dbDriver("PostgreSQL");con<-dbConnect(driver,user="vsyverson",password="gymnogyps",dbname="rlb")
test<-fetch(dbSendQuery(con,"select * from helloworld;"))
test;rm(test)
```
The previous line should return the cheerful contents of the table "helloworld" from database "rlb", confirming that the connection is good. 
Tables in the database were assembled and cleaned by the scripts in "database-cleaning.R" in this folder, not in this notebook. 

---
## 14C ages

The following code pulls the 14C ages from the database, converts pit number to a factor, removes NAs, and runs rcarbon::calibrate(), returning the data formatted as (1) a list "pit.ages.cal" with age calibrations grouped by pit number and (2) a data frame "all.ages" containing calibrations for all the dates in the input table.
```{r}
ages<-dbGetQuery(con,"select * from c14samples;")
ages$pit<-factor(ages$pit)
ages$uncal14cage<-as.numeric(ages$uncal14cage);ages$uncal14cerror=as.numeric(ages$uncal14cerror)
ages.clean<-ages
calpits<-split(ages.clean,ages.clean$pit)
pit.ages.cal<-lapply(calpits,calibrate.pit.ages)
all.ages<-calibrate(x=as.numeric(ages.clean$uncal14cage),errors=as.numeric(ages.clean$uncal14cerror))
all.ages<-calibrate.pit.ages(ages.clean)
```
Summarize the calibrated ages using Bayesian CKDE algorithm of Brown (2017). This is annoying because you have to call the objects by name and not index for some reason.

```{r}
pit.randomdates<-sampleDates(pit.ages.cal$`23`)
pit.kde<-ckde(pit.randomdates,timeRange=c(100000,1000),bw=100)

```

---
## Specimen data

The table "species" contains all the species from RLB and associated papers with a list of the elements of each. Query it as follows:
```{r}
test<-dbGetQuery(con,"select * from species where 'M1'=any(Elements)")
test;rm(test)
```
The table "tablenames" contains the full names for all the individual data tables, assembled automatically when the database is populated. Check names by querying it:
```{r}
test<-dbGetQuery(con,"SELECT * FROM tablenames WHERE tablenames LIKE '%buteo%';")
test;rm(test)
```
You can call tables using these names (but unlist them first!).
```{r}
tables.to.get<-unlist(dbGetQuery(con,"SELECT * FROM tablenames WHERE tablenames LIKE '%astragalus%';"))
astragalus.tables<-setNames(as.list(c(1:length(tables.to.get))),tables.to.get)
for (i in 1:length(tables.to.get)){
  query<-paste0("SELECT * FROM ",tables.to.get[i],";",collapse = "")
  astragalus.tables[[i]]<-dbGetQuery(con,query)
}
summary(astragalus.tables)
```


